---
title: "K-means models for MAPseq"
author: "Huizi Zhang"
date: "2025-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi=300,
  comment = "#>",
  fig.path = "man/figures/README-"
)
knitr::knit_hooks$set(imgcenter = function(before, options, envir){
  if (before) {
    HTML("<p align='center'>")
  } else {
    HTML("</p>")
  }
})
```

First, we load the data and necessary functions for MCMC sampling.

```{r package}
# load required packages
library(ggplot2)
library(HBMAP)
```


```{r data, warning=FALSE}
data("data_Hans")

M <- length(data_Hans)
C <- sapply(1:M, function(m) ncol(data_Hans[[m]]))
R <- dim(data_Hans[[1]])[1]

# Mouse index
mouse.index <- c(rep(1, C[1]),
                 rep(2, C[2]),
                 rep(3, C[3]),
                 rep(4, C[4]))
```

Implement the k-means model with varying choices of data transformation through the argument `transformation`: row-sum normalization, log of row-sum normalization, max-row normalization, cosine normalization.

```{r kmeans}
set.seed(43)
Z <- k_means_axon(Y = data_Hans, k = 20, transformation = 'log row-sum', restart = 20, iter.max = 100)

# make sure cluster labels are not continuous
Z_unlist <- unlist(Z)
uniq_cl <- unique(Z_unlist)
J <- length(uniq_cl)
Z_updated <- rep(0,sum(C))
for(j in 1:J){
  Z_updated[Z_unlist==uniq_cl[j]]=j
}
# rearrange in a list
# Cumulative number of neurons in a vector form
C_cumsum <- c(0, cumsum(C))
Z <- lapply(1:M,
            function(m) Z_updated[(C_cumsum[m]+1):C_cumsum[m+1]])
rm(Z_updated, C_cumsum, uniq_cl, Z_unlist, j, J)
```

Next, we reorder the clusters based on the average empirical strengths within each cluster and label each cluster by thresholding the average empirical strengths.

```{r reorder-kmeans-cluster}
# compute average empirical strengths for each motif
average_ps <- avg_ps(Y = data_Hans, Z = Z)
# reorder
kmeans_reorder <- reorder_cluster(proj_strength = average_ps, Z = Z)
# allocations after reordering
kmeans_Z_reorder <- kmeans_reorder$Z
# label clusters by projection regions with average_ps > thresh
thresh <- 0.05
cluster.labels <- sapply(1:length(unique(unlist(kmeans_Z_reorder))),function(j) {
    paste(j, ':', paste(rownames(data_Hans[[1]])[average_ps[j,] >= thresh], collapse = ','))
  })
```

Below we summarize cluster size, visualize neuron projection strengths within each cluster.

```{r cluster summary and line plots, message=FALSE, out.width='70%', fig.align='center'}
# you can choose your preferred color palette (change 'col' argument)
## ------ cluster size: number of neurons in each cluster, colored by group (mouse/injection site) ---------
opt.clustering.frequency(clustering = kmeans_Z_reorder, group.index = mouse.index, group.name = 'mouse', 
                         title = 'Cluster size')

## ---------- Heatmap of empirical projection strength of neurons in each cluster, colored by group (mouse/injection site) ----------
heatmap_ps(Y = data_Hans, Z = kmeans_Z_reorder, regions.name = rownames(data_Hans[[1]]), 
           group.index = mouse.index, group.name = 'mouse',
           cluster.index = 1:length(unique(unlist(kmeans_Z_reorder))), title = '')


## --------- Line plot for of empirical projection strengths within each cluster ------------------------
plot_empirical_ps(Y = data_Hans, Z = kmeans_Z_reorder, 
                  cluster.labels = cluster.labels,
                  regions.name = rownames(data_Hans[[1]]),
                  group.index = mouse.index, group.name = 'mouse',
                  cluster.index = 1:length(unique(unlist(kmeans_Z_reorder))),
                  title = 'Empirical projection strength', facet_ncol = 5)

```




